---
title: "615 Final Project--MBTA EDA"
author: "Yaquan Yang"
date: "2022-12-12"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
```{r message=F,warning=F,echo=FALSE}
library(data.table)
library(ggplot2)
library(stringr)
library(RColorBrewer)
library(flextable)
library(dplyr)
library(scales)
options(width=90)
```

# Subway data


This analysis uses data from November 2021 to October 2022 for the Red line, Blue line and Green-B line to analyze.

Red Line: From Alewife in the northwest to Braintree in the southeast.
Blue Line: From Wonderland to Bowdoin via Airport and Downtown.
Green-B: From Government Center to Boston College, passes through Boston University and Fenway Park. 

These three subway lines contain several important transportation hubs in the Boston area that we believe are worth analyzing.

```{r}
data.files<-list.files('MetroData',pattern='csv$',full=T)
data.files

dat<-lapply(setNames(,data.files),function(x)
{
   fread(x)->inter
   inter[,fromFile:=x]
   inter[,day:=mday(service_date)]
   inter<-inter[day %in% 1:7]
   inter<-inter[route_id %in% c('Red','Blue','Green-B')]
})


rbindlist(dat)->dat
dat[,startTime:=as.ITime(service_date)+start_time_sec]
dat[,endTime:=as.ITime(service_date)+end_time_sec]


fread('stops.txt')->sites
setNames(sites[,stop_name],sites[,stop_id])->sites.name

dat[,from_stop_name:=sites.name[as.character(from_stop_id)]]
dat[,to_stop_name:=sites.name[as.character(to_stop_id)]]

#fwrite(dat,file='LR.csv',sep=",")
#save(dat,sites,sites.name,file='LR.RData')
```

# Subway EDA 

## Total number of inbound and outbound trains for different lines (Red,Blue,Green-B) during the year

```{r}
pie.data<-dat[,.N,route_id]
pie.data[,percent:=percent(N/sum(N),0.01)]
pie.data[,label:=paste0(route_id," (",comma(N),"; ",percent,")")]


pie(pie.data$N,
    pie.data$label,
    col=c('darkolivegreen','olivedrab3','khaki'), main= "Total number of inbound and outbound trains")
```

From the above graph we can see that the traffic flow on the red and green-B lines is much higher than the blue line, which means that the red and green-B lines are busier than the blue line most of the time.


## Total number of inbound and outbound trains at different stations

```{r}
fromVolume<-dat[,.N,from_stop_id][order(-N)]
names(fromVolume)<-c('stop_id','from_count')
toVolume<-dat[,.N,to_stop_id][order(-N)]
names(toVolume)<-c('stop_id','to_count')
siteVolume<-fromVolume[toVolume,on=.(stop_id)]
siteVolume[,stop_name:=sites.name[as.character(stop_id)]]
siteVolume[,total_count:=from_count+to_count]
siteVolume<-siteVolume[order(-total_count)]
siteVolume<-siteVolume[,.(stop_id,stop_name,from_count,to_count,total_count)]


siteVolume %>%
head(15) %>%
    flextable() %>% 
    width(j=2,width=2) %>%
    theme_vanilla() %>%
	set_caption('Top 15 of the heaviest traffic STOPs according total count')

site.plot.data<-siteVolume %>% head(15)
site.plot.data[,stop_name:=ordered(stop_name)]
site.plot.data<-site.plot.data[,.(stop_name,from_count,to_count)]
site.plot.data<-melt(site.plot.data,id.var='stop_name')


ggplot(site.plot.data,aes(x=stop_name,y=value,fill=gsub('_count','',variable)))+
geom_bar(stat='identity')+
theme_bw()+
guides(x=guide_axis(angle=60))+
labs(x='Stop Name',
     y='Total Count',fill='',title = "Top 15 of the heaviest traffic STOPs according total count")

```

We can see that the busiest stops are "Alewife", "South Station" and "Downtown Crossing"

## 

```{r}
# tra<-dat[,.(Traffic=.N),.(from_stop_id,to_stop_id)]
# tra[,from_stop_id:=as.character(from_stop_id)]
# tra[,to_stop_id:=as.character(to_stop_id)]
# 
# 
# to.anno<-sites[,.(to_stop_id=stop_id,to_stop_lat=stop_lat, to_stop_lon=stop_lon,to_stop_name=stop_name)]
# tra<-to.anno[tra,on=.(to_stop_id)]
# from.anno<-sites[,.(from_stop_id=stop_id,from_stop_lat=stop_lat,  from_stop_lon=stop_lon,from_stop_name=stop_name)]
# tra<-from.anno[tra,on=.(from_stop_id)]
# tra<-tra[order(-Traffic)]
# 
# 
# tra[,.(from_stop_name,to_stop_name,Traffic)] %>% 
#     head(15) %>%
#     flextable() %>% 
#     width(j=1:2,width=2) %>%
#     theme_vanilla() %>%
# 	set_caption('Top 15 of the heaviest traffic between two subway stations')
# 
# 
# ggplot(tra,aes(x=from_stop_lon,xend=to_stop_lon,y=from_stop_lat,yend=to_stop_lat,colour=Traffic))+
#    geom_segment()+
#    geom_point(colour='black')+
#    scale_colour_gradientn(colours=c(rev(brewer.pal(5,'YlGn')),brewer.pal(5,'YlOrRd')))+
#    coord_sf()+
#    theme_minimal()+
#    labs(x='Longitude',y='Latitude',
#         title='Distribution of stops and the traffic flow between them')
```



## Distribution of subway traffic and time

### Distribution of subway traffic within 24-hours

```{r}
dat[,startHour:=hour(startTime)]
hour.data<-dat[,.(count=.N),startHour][order(startHour)]
hour.data<-hour.data[data.table(startHour=0:23),on=.(startHour)]

flextable(hour.data) %>% 
    width(j=2,width=2) %>%
    theme_vanilla() %>%
	set_caption('Distribution of subway traffic within 24-hours')


ggplot(hour.data,aes(x=startHour,y=count))+
    geom_bar(stat='identity',fill='orange')+
    theme_minimal()+
    scale_x_continuous(breaks=0:23)+
    theme(panel.grid.minor=element_blank())+
    labs(x='Start Hour',
     y='Total Count',fill='',title = "Distribution of subway traffic within 24-hours")
```

The morning peak of the subway is at 7:00-9:00 a.m., the evening peak is at 4:00-7:00 p.m., and almost all trains stop running from 1:00 a.m. to 4:00 a.m.

### Weekday

```{r}
weekdays.name<-strsplit('Sun,Mon,Tue,Wed,Thu,Fri,Sat',',')[[1]]

dat[,weekday:=wday(service_date)]
dat[,weekDay:=weekdays.name[as.integer(weekday)]]
dat[,weekDay:=ordered(weekDay,weekdays.name)]

week.data<-dat[,.(count=.N),weekDay][order(weekDay)]

flextable(week.data) %>% 
    width(j=2,width=2) %>%
    theme_vanilla() %>%
	set_caption('Distribution of subway traffic in 7 weekdays')


ggplot(week.data,aes(y=count,x=weekDay))+
  geom_bar(stat='identity',fill='darkseagreen3',width=0.6)+
  theme_minimal()+
    labs(x='Weekday',
     y='Total Count',fill='',title = "Distribution of subway traffic in 7 weekdays")
```


## Month

```{r}
dat[,Month:=month(service_date)]
dat[,MonthName:=month.abb[Month]]
dat[,MonthName:=ordered(MonthName,month.abb)]

month.data<-dat[,.(count=.N),MonthName][order(MonthName)]

flextable(month.data) %>% 
    width(j=2,width=2) %>%
    theme_vanilla() %>%
	set_caption('Distribution of subway traffic in different month')


ggplot(month.data,aes(y=count,x=MonthName))+
  geom_bar(stat='identity',fill='darkseagreen3',width=0.6)+
  theme_minimal()+
  labs(x='Month',
     y='Total Count',fill='',title = "Distribution of subway traffic in different month")
```


## Distribution of travel_time

```{r}
ggplot(dat,aes(travel_time_sec/60))+
geom_histogram(bins=30,fill='pink1',colour='pink4')+
theme_bw()+
    labs(x='Travel time(min)',
     y='Total Count',fill='',title = "Distribution of travel time")
```

From the above chart, we can see that most of the travel_time are located within 50 minutes, let's look at the distribution of travel_time < 50 mins:

```{r}
ggplot(dat[travel_time_sec<=3000],aes(travel_time_sec/60))+
geom_histogram(bins=30,fill='pink1',colour='pink4')+
theme_bw()+
        labs(x='Travel time(min)',
     y='Total Count',fill='',title = "Distribution of travel time (<50 mins)")
```

Distribution of travel_time for 7 days in a week:

```{r fig.height=7.8,message=F,warning=F}
library(ggridges)

ggplot(dat[travel_time_sec<=3000],aes(x=travel_time_sec/60,y=weekDay,fill=weekDay))+
   geom_density_ridges(alpha=0.7)+
   theme_bw()+
    labs(x='Travel time(min)',
     y='Weekday',fill='',title = "Distribution of travel time for 7 days in a week")

```

The results show that the distribution is similar.


# Bus Data

We choose three routes: No.8, No.56 and No.71 for analysis

```{r}
bus.data.files<-list.files('BusData',pattern='csv$',full=T)

bus<-lapply(setNames(,bus.data.files),function(x)
{
fread(x)->inter
})

bus<-rbindlist(bus)
bus<-bus[route_id %in% c('08','57','71')]
bus<-bus[!point_type %in% c('Pullout','Pullback')]

bus[,day:=mday(service_date)]
bus<-bus[day %in% 1:7]

bus<-bus[,.(service_date,route_id,stop_id,point_type,scheduled,actual,day)]
bus<-na.omit(bus)

bus[,timeDiff:=as.numeric(actual-scheduled)]
bus[,sum(abs(timeDiff)<=1800)/.N] 

bus<-bus[abs(timeDiff)<=1800]

fread('stops.txt')->sites
setNames(sites[,stop_name],sites[,stop_id])->sites.name

bus[,stop_name:=sites.name[as.character(stop_id)]]

weekdays.name<-strsplit('Sun,Mon,Tue,Wed,Thu,Fri,Sat',',')[[1]]
bus[,weekday:=wday(service_date)]
bus[,weekDay:=weekdays.name[as.integer(weekday)]]
bus[,weekDay:=ordered(weekDay,weekdays.name)]


#save(bus,file='bus.RData')
#fwrite(bus,file='bus.csv',sep=",")
```

# Bus EDA

```{r message=F,warning=F}
library(data.table)
library(ggplot2)
library(stringr)
library(RColorBrewer)
library(flextable)
library(dplyr)
library(scales)
library(ggridges)

bus[,scheduledHour:=as.character(hour(scheduled))]
bus[,scheduledHour:=ordered(scheduledHour,0:23)]
bus[,type:=fcase(timeDiff==0,'Intime',timeDiff>0,'Delay',timeDiff<0,'Advance')]

head(bus)
```


## The distribution of Delay/Intime/Advance in the three routes


```{r fig.width=4.2}
plot1.data<-bus[,.(count=.N),.(type,route_id)]

ggplot(plot1.data,aes(x=route_id,y=count,fill=type))+
geom_bar(stat='identity',width=0.6)+
theme_bw()+
    labs(x='Route id',
     y='Total Count',fill='',title = "Bus arrival status")
```

## The distribution of Delay/Intime/Advance within 24 hours

```{r fig.width=9.3}
plot2.data<-bus[,.(count=.N),.(type,scheduledHour)]

ggplot(plot2.data,aes(x=scheduledHour,y=count,fill=type))+
geom_bar(stat='identity')+
theme_bw()+
    labs(x='Schedule Hour',
     y='Total Count',fill='',title = "Bus arrival status within 24 hours")
```

## Top 15 sites with the most delays

```{r}
bus[type=='Delay',.(sumDelayTime=sum(timeDiff)),.(stop_name)][order(-sumDelayTime)] %>% 
head(15) %>%
    flextable() %>% 
    width(j=1,width=4) %>%
    theme_vanilla() %>%
    set_caption('Top 15 sites with the most delays')
```

The stations with the most severe delays are "Tremont St @ Washington St", "Kenmore" and "Watertown Yard".

## Distribution of dalay time

```{r}
ggplot(bus,aes(y=point_type,x=timeDiff/60,fill=point_type))+
  geom_density_ridges(alpha=0.5)+
  labs(x='Actual-Scheduled(min)',
     y='Point type',fill='',title = "Distribution of dalay time")
```

From the above chart can see the Startpoint punctuality is high, the Midpoint has a high probability of delay, but the Endpoint, some instead arrived in time or even early.
And most of the bus are delayed within 10 minutes.

```{r}
ggplot(bus,aes(y=route_id,x=timeDiff/60,fill=route_id))+
  geom_density_ridges(alpha=0.5)+
  labs(x='Actual-Scheduled(min)',
     y='Route id',fill='',title = "Distribution of dalay time for different routes")
```

As you can see from the chart above, route 71 is more punctual and 08 and 57 are more delayed.

## Distribution of dalay time for 7 days in a week

```{r}
ggplot(bus,aes(x=weekDay,y=timeDiff/60,colour=weekDay,fill = after_scale(alpha(colour, 0.4))))+
   geom_boxplot()+
   theme_bw()+
   scale_colour_hue()+
  labs(x='Actual-Scheduled(min)',
     y='WeekDay',fill='',title = "Distribution of dalay time for 7 days in a week")
```

The delay is more severe on weekdays than on Saturdays and Sundays.

## Distribution of dalay time within 24 hours

```{r fig.width=9.3}
ggplot(bus,aes(x=scheduledHour,y=timeDiff/60,colour=scheduledHour,fill = after_scale(alpha(colour, 0.4))))+
   geom_boxplot()+
   theme_bw()+
   scale_colour_hue()+
   theme(legend.position='none')+
  labs(x='Actual-Scheduled(min)',
     y='Scheduled Hour',fill='',title = "Distribution of dalay time within 24 hours")
```


The delay varies from time to time, but during commuting hours, the delay is more serious. And in the morning and evening rush hour delays can exceed ten minutes.




